---
title: "Data in R"
author: "Haley Carter"
date: "2023-01-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Working with data in R requires getting data into your R environment. At the end of the Basics tutorial we put together a dataframe of information we made up about abandoned lots. For small datasets, manual entry like that might be worth doing so let's revisit those functions.

## Manual entry

* c() combines items into a list or vector
* rbind() combines vectors into a dataframe as rows
* cbind() combines vectors into a dataframe as columns

### Exercise 1: make up some data!

```{r}
# make up three lists you can combine in different ways
# need some inspiration? how about...
# three lists of the age, favorite color, and birthday of people you know. do these make sense to combine as rows or columns?
# a list of books, a list of their authors, and a list of whether or not you like them
# do these make sense to combine as rows or columns?
```


Note that my suggestions are within the code block but they don't run as lines of code. This is because they start with # which makes them 'comments'. Comments are useful for putting notes in your code. 

## Reading in data files from your computer

Odds are, recording your data in R is not the best way forward for real research projects. Many people enter their data in Excel. We can read those files into R in a couple ways.

### Files types from Excel

You can save different files from Excel. Most newer versions default to xlsx and older versions to xls. Either of these can be read into R without conversion, but it does require a function that is not included in the basic functions of R. That will be covered a little later. 

A great option is to save files as comma separated values or a .csv file. This is possible through Excel using "Save as..." but it will only save one sheet of a workbook so if you're using multiple tabs they'll need to be saved separately. 

Let's read in...

```{r}
#data_csv <- read.csv()
```


### Other packages

```{r}

```

